The Hot Hand Fallacy in Call of Duty

### 2022-08-22 paper/paper.Rmd
one big self review

---
title: "The Hot Hand in Call of Duty Search and Destroy"
authors:
  - name: Tony ElHabr
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
  - name: Andrew ElHabr
    email: andrewelhabr@gatech.edu
    affiliation: Georgia Institute of Technology
    department: School of Industrial and Systems Engineering (ISyE)
    location: Atlanta, Georgia, USA
abstract: |
  Our research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD). First, we find evidence supprting the hypothesis that round win probability can be modeled as a constant across the series, although not at the naive 50%. Second, we examine post-streak round win probability, given the series outcome. We find no evidence that post-streak win rate is significantly different than expected for combinations of streak length, series length, and series winner, suggesting that the perception of momentum at the team level is just that, perception.
bibliography: references.bib
keywords:
  - esports
  - Call of Duty
  - hot hand
  - runs test
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

@@

Call of Duty (CoD), first released in 2003, is one of the most popular first-person shooter (FPS) video game franchises of all-time. The most popular mode in the competitive scene is "Search and Destroy" (SnD).[^1] SnD is a one-sided game mode in which one team, the offensive side, tries to destroy one of two designated bomb sites on the map.

@@

In professional CoD SnD, the two teams take turns playing offense and defense every round. They must win six rounds to win the series.[^2] A round can end in one of five ways:

[^2]: A maximum of 11 even rounds can be played. There is no "sudden death" or "win by two" rule like there are for SnD equivalent in professional Counter-Strike and Valorant matches.

1.  One team eliminates all members of the other team prior to a bomb plant. (Eliminating team wins.)
2.  The offensive team eliminates all members of the defensive team after a bomb plant.[^3] (Offense wins.)
3.  The defensive team defuses the bomb after a bomb plant.[^4] (Defense wins.)
4.  The offensive team does not make a plant by the time the round timer ends. (Defense wins.)

[^3]:
    -   The bomb can be picked up by any member of the offensive team.
    -   The bomb carrier is not obstructed at all by carrying the bomb (i.e. movement is the same, weapon usage is the same).
    -   The defense does not get any visual indication for who is carrying the bomb.
    -   A bomb plant takes five seconds. The timer resets if the player stops planting site prior to completing it.
    -   A bomb defuse takes seven seconds. The timer resets if the player "drops" the bomb.
    -   The bomb takes 45 seconds to defuse after being planted.

[^4]: Often the defensive team will try to eliminate all team members prior to making the defuse, but in some cases, they may try to "ninja" defuse.

We adopt the terminology "series" to refer to what CoD SnD players typically call a "match", so as to emulate the terminology of playoff series in professional leagues like the National Basketball Association, National Hockey League, and Major League Baseball. A "game" or a "match" in such leagues is analogous to a "round" of CoD SnD. Further, it is analogous to a trial in statistical analysis of Bernoulli events.

SnD is not the only game type played in competitive CoD. Teams play in a head-to-head, best-of-five (first-to-three wins) format, where SnD is always played as the second and fifth game mode (if a fifth variant is needed). The best-of-five across game types could also be called a "series", but since we analyze only the SnD game types, we refer to SnD games as series (of rounds).

Anecdotally, most observers swear by the existence of momentum in CoD SnD series, to the extent that vernacular has been developed to describe such phenomenon. People say that teams "ice up" when the team wins several consecutive rounds to come back to win a series, often culminating with a series-clinching round win.[^5] Further, viewers have come to embrace the "5-3" phenomenon, where teams win three consecutive rounds after facing a 5-3 deficit to win 6-5. There is even a term for the rare 0-5 comeback---a "full sail".

[^5]: This vernacular is not specific to CoD SnD in specific, and is not only used in the contexts of streaks.

@@

CoD has roughly gone through three eras of professional gaming: (1) Major League Gaming (MLG) tournaments prior to 2016; (2) the CoD World League (CWL), initiated in 2016; and (3) the 12-franchise CoD League (CDL), operating since 2020. The CDL has completed three year-long "seasons" as of August 2022.[^6]

[^6]: CoD is fairly unique compared to other esports in that it runs on an annual lifecycle (released coming in the late fall), where a new game is published every year under the same title. Each new game bears resemblance to past ones, often introducing relatively small variations ("improvements") to graphics, game modes, and other facets of gameplay. During the CDL era, the games released have been Modern Warfare (2020), Cold War (2021) and Vanguard (2022).

The data set consists of all SnD matches played in major tournaments and qualifiers during the CDL era, totaling 7,792 rounds across 852 series.[^7] Data was collected in spreadsheets by community member "IOUTurtle".[^8]

[^7]: 288 of the series occur in major tournaments, which are considered to be more "competitive" than the qualifiers since they are played in person (COVID-permitting), whereas the qualifiers are played online.

[^8]: Raw data: <https://linktr.ee/CDLArchive>. Author: <https://twitter.com/IOUTurtle>. Processed data can be found at <https://github.com/tonyelhabr/fps_round_games/blob/master/data/cod_rounds.csv>.

The observed offensive round win percentage across all rounds is $\tau_O$ = 47.8%.[^9] Table \ref{tbl:cod-o-win-prop-by-series-state} shows round win percentages by series "state" (i.e. the number of round wins by each team prior to an upcoming round). Offensive round win rate is not quite constant, although never veers more than 10% from this overall average.

[^9]: Offensive round win percentage has been nearly constant across the three games during the CDL era: 1. 47.2% in MW (2020) 2. 47.9% in Cold War (2021) 3. 48.1% in Vanguard (2022)

@@

```{=tex}
\begin{longtable}{crrrrrr}
\caption{Offensive round win rates for the upcoming round, given both the offensive and defensive team's prior number of round wins. Numbers in parentheses are sample sizes.}\label{tbl:cod-o-win-prop-by-series-state} \\
\toprule
& \multicolumn{6}{c}{Offense round wins} \\ 
\cmidrule(lr){2-7}
Defense round wins & 0 & 1 & 2 & 3 & 4 & 5 \\ 
\midrule
0 & 47.8\%
(852) & 46.6\%
(408) & 43.1\%
(216) & 43.5\%
(115) & 43.3\%
(67) & 40.5\%
(37) \\ 
1 & 48.6\%
(444) & 49.3\%
(418) & 51.5\%
(309) & 43.4\%
(205) & 43.3\%
(120) & 39.4\%
(99) \\ 
2 & 52.8\%
(218) & 48.9\%
(305) & 48.9\%
(315) & 46.6\%
(262) & 48.7\%
(189) & 42.1\%
(133) \\ 
3 & 54.5\%
(123) & 46.0\%
(200) & 49.6\%
(250) & 45.6\%
(248) & 44.4\%
(214) & 44.8\%
(174) \\ 
4 & 56.9\%
(65) & 54.5\%
(145) & 47.2\%
(193) & 44.7\%
(228) & 55.2\%
(221) & 50.5\%
(208) \\ 
5 & 47.4\%
(38) & 49.4\%
(83) & 47.1\%
(136) & 50.9\%
(175) & 45.2\%
(177) & 46.0\%
(202) \\ 
\bottomrule
\end{longtable}
```
On the extremes, 6-0 sweeps make up 4.7% of all series, while 6-5 series make up 23.7 of series.

At tournaments, higher-seeded ("better") teams choose whether they want to start on offense in the SnD series played as the second game or the fifth game, and their opponent is assigned to play offense in the first round of the SnD slot not chosen.[^10][^11] [@codrules] With online qualifiers, teams all play each other twice throughout the season. Each team gets to play the higher-seeded role once in their matchups, despite their win-loss records. Overall, we might say that bias due to stronger teams playing defense (slightly advantageous) more often is very minimal.

[^10]: "Hardpoint" has been played as the first and fourth games in a matchup; the variant for the third game was Domination in the 2020 season, and Control in the 2021 and 2022 seasons.

[^11]: Anecdotally, the higher-seeded teams tend to choose to play defense in the first round of the SnD series played as the second game, although this choice is not consistent across teams, or with the same team over time.

@@

There have been a handful of studies of the distribution of games played in a series of a professional sport. Most assume a constant probability $\phi$ of a given team winning a game in the series, regardless of the series state. Mosteller [-@mosteller1952] observed that the American League had dominated the National League in Major League Baseball's (MLB) World Series matchups, implying that games should not be modeled with a constant (null) $\phi_0 = 0.5$. Mosteller proposed three approaches for identifying the optimal constant probability value of the stronger team in the World Series, finding $\phi \approx 0.65$ in each case: (1) solving for $\phi$ from the observed average number of games won by the loser of the series, which he called the "method of moments" approach; (2) maximizing the likelihood that the sample would have been drawn from a population in which the probability of a team winning a game is constant across the series (i.e. maximum likelihood), and (3) minimizing the chi-square goodness of fit statistic $\chi^2$ for $\phi$.

Chance [-@chance2020] re-examines the constant probability notion in Major League Baseball's (MLB) World Series (1923--2018), the National Basketball Association's (NBA) Finals (1951--2018), and the National Hockey League's (NHL) Stanley Cup (1939--2018). Chance finds strong evidence against the null hypothesis of $\phi_0 = 0.5$ in the MLB and NHL championship series when applying Mosteller's first and second methods. Chance goes on to outline a conditional probability framework (likelihood of winning a game given the series state) which can exactly explain the distribution of the number of games played.

Momentum, one of most discussed topics in sports analytics, goes hand-in-hand with a discussion of the nature of series outcomes.[^12] Two opposing fallacies are observed in the context of momentum: the "gambler's fallacy" (negative recency) and the "hot hand fallacy" (positive recency). Per Ayton et al. [-@ayton2004], negative recency is "the belief that, for random events, runs of a particular outcome ... will be balanced by a tendency for the opposite outcome", while positive recency is the expectation of observing future results that match recent results.

[^12]: We often use use "streaks" and momentum interchangeably, but as [@steeger2021] note, momentum implies dependence between events, whereas streaking does not.

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985], henceforth GVT, do not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018], henceforth MS, provide a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy. Specifically, MS say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to GVT's data, MS come to the opposite conclusions as GVT.

Other research has borrowed a common approach from the field of quality control, looking at unlikely sequences of events with the Wald-Wolfowitz runs test [@peel2015], [@steeger2021]. Peel and Clauset find no evidence for unlikely sequences of scoring events in the NHL, College Football (CFB), and National Football League (NFL), but do in the NBA. As a check on their entropy approach to momentum identification, Steeger et al. found three NHL teams who sequence of wins in the 2018-2019 regular season violated the Wald-Wolfowitz null hypothesis, although these teams did not match the teams identified with their primary approach.

@@

Despite the plethora of existing research on games played in a series and momentum in sports, these topics have yet to be investigated heavily, if at all, in esports. Work has been done to examine intra-round win probability in other FPS titles such as Counter-Strike [@xenopoulos2022] and Valorant [@derover2021], both of which are round-based like CoD SnD. However, considering Counter-Strike and Valorant specifically, research on round-level trends seems non-existant, perhaps for one of many reasons:

1.  Both have economic aspects that can create clear advantages for one side in a given round, given how prior rounds played out. CoD has no such equivalent, except for perhaps "kill streaks", which infrequently play a role.
2.  Teams play either offense or defense for many consecutive rounds. (The former has a 14-14-1 format and the latter uses a 12-12-1 format.) On the other hand, teams in CoD SnD rotate roles every round, analogous to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series for professional sports like the MLB, NBA, and NHL.[^13] While theoretically one might be able to account for any kind of format, such as a 5-5-1, the rotation of team sides every round is convenient for convincing ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli trials.
3.  Both Counter-Strike and Valorant have overtime rules---team must win by two rounds---which can make end-of-series sequences difficult to model. CoD SnD does not have overtime rules.

[^13]: 1-1-1-1-1-1-1 is not used today in these leagues, but it was at least once in each league.

To our knowledge, there is no existing public statistical research on CoD SnD format, beyond descriptive analysis on social media. While intra-round trends may be more directly applicable to teams looking for an advantage on their competition, broader investigation on a common concept like the hot-hand phenomenon in a new field---particularly one that is less subjected to factors that may be difficult to control for, e.g. weather---should be useful as a precedent or supplement for future researchers.

@@

In a best-of-$s$ format, assuming a constant round win probability $\phi$ ($0 \leq \phi \leq 1$) for one team[^14], the probability that a series ends in $r$ rounds ($r \leq s$), $\hat{\Phi}(r)$ is

[^14]: If $\phi > 0.5$, we can say that this is the stronger team.

```{=tex}
\begin{equation}\label{eq:m}
m = \frac{s + 1}{2}
\end{equation}
```
```{=tex}
\begin{equation}\label{eq:series-length}
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(s - r)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
\end{equation}
```
$s = 11$ rounds for CoD SnD.[^15] For example, assuming $\phi = 0.5$, the probability of a series ending in 9 rounds in CoD is

[^15]: $\phi$ and other symbols are selected so as to reserve symbols like $p$ for usage in other contexts without causing confusion for the reader. Upper-case symbol, e.g. $\Phi$, are consistently used in this paper to represent proportions, while lower-case symbols are used to represent Bernoulli trials success probability. "Hats", e.g. $\hat{\Phi}$, are used to convey expectations, while bare symbols symbols convey observational data.

@@

```{=tex}
\begin{equation}\label{eq:chi-squ}
\chi^2 = \sum^{r \in R} \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)}.
\end{equation}
```
in which $R = [6, 7, 8, 9, 10, 11]$ for CoD SnD.

@@

```{=tex}
\begin{equation}\label{eq:pwr}
p_0(\text{win} | r) = p^{+|r}_0 = \begin{cases} 
\frac{m}{r} & \text{team wins series} \\ 
\frac{r - m}{r} & \text{team loses series}  
\end{cases}
\end{equation}
```
using $m$ from Equation \ref{eq:m}.

As shown by MS with their Theorem 1, we should expect the proportion of rounds wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{+|k,r}_{MS}$, to be strictly less than $p^{+|r}$.[^16] Unfortunately, there does not exist a closed form representation for the expected value of $\hat{P}^{+|k,r}_{MS}$ for $k > 1$. Nonetheless, one may recursively build a collection of dictionaries to compute the expectation analytically, or, more conveniently, one may run simulations to estimate the expected value, as We choose to do.[^17]

[^16]: See Appendix E.1 of MS for the proof.

[^17]: We have adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

Although it would be convenient to emulate the hypothesis testing for difference of proportions performed by MS (and GVT), our context is fundamentally different from that of MS, who focus on longitudinal data in controlled settings.[^18] The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series. The Bernoulli trial success probability, i.e. a round win in CoD SnD, is not independent of the opponent. Consequently, a statistical test of the difference in $\hat{P}^{+|k,r}$ and $\hat{P}^{-|kr}$, as performed by MS to evaluate their hypothesis regarding post-streak success rates, is not completely appropriate, although useful as a reference.

[^18]: GVT also performs statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team.

@@

```{=tex}
\begin{equation}\label{eq:pwkr}
p_0(\text{win} | k, r) = p^{+|k,r}_0 = \begin{cases}
\frac{m - k}{r - k} & \text{team wins series} \\
\frac{s - m - k}{r - k} & \text{team loses series}
\end{cases}
\end{equation}
```
using $m$ from Equation \ref{eq:m}.

We can perform a binomial test to evaluate the null hypothesis $\omega = \omega_0$ for the observed probability of success $\omega$ and a user-specified $\omega_0$, where $0 \leq \omega_0 \leq 1$.[^19] If there are $r^+$ observed successes in a sample of $r$ trials and we expect that there should be $r * \omega_0$, the probability of arriving at this expected number of successes is

[^19]: $\omega$ and other symbols are chosen so as to not conflict with symbols already used elsewhere for in other equations, either generic or for CoD SnD specifically.

```{=tex}
\begin{equation}\label{eq:binom}
\Pr(r^w) = {\binom {r}{r^+}} \omega^{r^+}(1-\omega)^{r-r^+}.
\end{equation}
```
Treating the notional proportion $p^{+|k,r}_0$ as the null $\omega_0$ and plugging in the observed proportion $P^{+|k,r}$ for $\omega$ in Equation \ref{eq:binom} (treating proportions as probabilities), we can evaluate the null hypothesis that the observed probability is equal to the notional probability. If we can reject this null hypothesis, then we can consider team momentum, represented by post-streak success, plausible.

One can perform the same binomial test for the MS's streak-selection-adjusted proportion, $\hat{P}^{+|k,r}_{MS}$. FOOTY: Given the caveats mentioned before, the results should be heeded with caution.

We can further decompose Equation \ref{eq:pwkr} by the round $i$ ($i \leq r$) in which the streak of length $k$ carries into.

>>>>

---
title: "The Hot Hand Fallacy in Call of Duty"
authors:
  - name: Tony ElHabr
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
  - name: Andrew ElHabr
    email: andrewelhabr@gatech.edu
    affiliation: Georgia Institute of Technology
    department: School of Industrial and Systems Engineering (ISyE)
    location: Atlanta, Georgia, USA
abstract: |
  Our research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD). First, we find evidence suppOrting the hypothesis that round win probability can be modeled as a constant across rounds in the series, although not at the naive 50%. Second, we examine post-streak round win probability, given the series outcome. Given streak length, series length, and series winner, we find no evidence that post-streak win rate is significantly different than expected, suggesting that the perception of momentum at the team level is deceiving. Wald-Wolfowitz run tests also fail to provide evidence for the hot hand phenomenon.
bibliography: references.bib
keywords:
  - esports
  - Call of Duty
  - hot hand
  - runs test
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

@@

Call of Duty (CoD), first released in 2003, is one of the most popular first-person shooter (FPS) video game franchises of all-time. The most popular mode in the competitive scene is "Search and Destroy" (SnD), in which one team tries to destroy one of two designated bomb sites on the map, defended by the opposing team.[^1]

@@

In professional CoD SnD, the two teams[^2] take turns playing offense and defense every round. The first to six round wins (best-of-11 format) is declared the series winner.[^3] A round can end in one of four ways:

[^2]: In the 2020 season, teams played with five players each; in the 2021 and 2022 seasons, teams played with four players each.

[^3]: Each round has a two-minute time limit, not counting the potential extension granted if a bomb is planted.

1.  One team eliminates all members of the other team prior to a bomb plant. (Eliminating team wins.)
2.  The offensive team eliminates all members of the defensive team after a bomb plant.[^4] (Offense wins.)
3.  The defense defuses the bomb after a bomb plant.[^5] (Defense wins.)
4.  The offense does not make a plant by the time the round timer ends. (Defense wins.)

[^4]:
    -   The bomb can be picked up by any member of the offensive team.
    -   The bomb carrier is not obstructed by carrying the bomb (i.e. movement is the same, weapon usage is the same, etc.).
    -   The defense does not get any visual indication for who is carrying the bomb.
    -   A bomb plant takes five seconds. The timer resets if the player stops planting site prior to completing it.
    -   A bomb defuse takes seven seconds. The timer resets if the player "drops" the bomb.
    -   Once the bomb is planted, the round timer is reset to 45 seconds.

[^5]: Often the defensive team will try to eliminate all team members prior to making the defuse, but in some cases, they may try to "ninja" defuse.

We adopt the terminology "series" to refer to a single best-of-11 matchup, so as to mirror the terminology of playoff series in professional leagues like the National Basketball Association (NBA), National Hockey League (NHL), and Major League Baseball (MLB). A "game" or a "match" in such leagues is analogous to a "round" of CoD SnD for the purposes of this paper.[^6]

[^6]: Further, these are analogous to a trial in statistical analysis of Bernoulli events.

SnD is not the only game type played in competitive CoD. Teams play in a head-to-head, best-of-five format, where SnD is always played as the second and fifth game "maps", as most players refer to them (if a fifth map is needed). The best-of-five matchup could also be called a "series", but since we analyze only the SnD game types, we refer to SnD games as series.

@@

CoD has roughly gone through three eras of professional gaming: (1) Major League Gaming (MLG) tournaments prior to 2016; (2) the CoD World League (CWL), initiated in 2016; and (3) the 12-franchise CoD League (CDL), operating since 2020. The CDL has completed three year-long "seasons" as of August 2022.[^7]

[^7]: CoD is fairly unique compared to other esports in that it runs on an annual lifecycle, with releases coming in the late fall. A new game under the same brand---Call of Duty---is published every year by a rotating set of developers. Each new game bears resemblance to past ones, often introducing relatively small variations ("improvements") to graphics, game modes, and other facets of gameplay. During the CDL era, the games released have been Modern Warfare (2020), Cold War (2021) and Vanguard (2022).

The data set consists of all SnD matches played in major tournaments and qualifiers during the CDL era, totaling 7,792 rounds across 852 series.[^8] Data has been collected in spreadsheets by community member "IOUTurtle".[^9]

[^8]: 288 of the series occur in major tournaments, which are considered to be more "competitive" than the qualifiers since they are played in person (COVID-permitting), whereas the qualifiers are played online.

[^9]: Raw data: <https://linktr.ee/CDLArchive>. Author: <https://twitter.com/IOUTurtle>. Processed data can be found at <https://github.com/tonyelhabr/fps_round_games/blob/master/data/cod_rounds.csv>.

6-0 sweeps make up 4.7% of all series, while 6-5 series make up 23.7% of series. The observed offensive round win percentage across all rounds is $\tau_O$ = 47.8%.[^10] Table \ref{tbl:cod-o-win-prop-by-series-state} shows round win percentages by series "state" (i.e. the number of round wins by each team prior to an upcoming round). Offensive round win rate is not quite constant, although never veers more than 10% from $\tau_O$.

[^10]: Offensive round win percentage has been nearly constant across the three games during the CDL era: 1. 47.2% in MW (2020) 2. 47.9% in Cold War (2021) 3. 48.1% in Vanguard (2022)

@@

```{=tex}
\begin{longtable}{crrrrrr}
\caption{Offensive round win rates for the upcoming round, given both the offensive and defensive team's prior number of round wins. Numbers in parentheses are sample sizes.}\label{tbl:cod-o-win-prop-by-series-state} \\
\toprule
& \multicolumn{6}{c}{Offense round wins} \\ 
\cmidrule(lr){2-7}
Defense round wins & 0 & 1 & 2 & 3 & 4 & 5 \\ 
\midrule
0 & 47.8\%
(852) & 46.6\%
(408) & 43.1\%
(216) & 43.5\%
(115) & 43.3\%
(67) & 40.5\%
(37) \\ 
1 & 48.6\%
(444) & 49.3\%
(418) & 51.5\%
(309) & 43.4\%
(205) & 43.3\%
(120) & 39.4\%
(99) \\ 
2 & 52.8\%
(218) & 48.9\%
(305) & 48.9\%
(315) & 46.6\%
(262) & 48.7\%
(189) & 42.1\%
(133) \\ 
3 & 54.5\%
(123) & 46.0\%
(200) & 49.6\%
(250) & 45.6\%
(248) & 44.4\%
(214) & 44.8\%
(174) \\ 
4 & 56.9\%
(65) & 54.5\%
(145) & 47.2\%
(193) & 44.7\%
(228) & 55.2\%
(221) & 50.5\%
(208) \\ 
5 & 47.4\%
(38) & 49.4\%
(83) & 47.1\%
(136) & 50.9\%
(175) & 45.2\%
(177) & 46.0\%
(202) \\ 
\bottomrule
\end{longtable}
```
At tournaments, higher-seeded ("better") teams choose whether they want to start on offense in the SnD series played as the second or fifth map, and their opponent is assigned to start on offense in the SnD map not chosen.[^11][^12] Outside of tournaments, teams all play each other twice throughout the season, constituting qualifiers. Each team gets to play the higher-seeded role once in their head-to-head matchups with a given team, despite their win-loss records. Overall, we can argue that potential bias due to stronger teams playing defense---or, if offense, if offense happens to be generally advantageous for a given map---is very minimal.

[^11]: Hardpoint has been played as the first and fourth maps in a matchup. The variant for the third map was Domination in the 2020 season, and Control in the 2021 and 2022 seasons.

[^12]: Anecdotally, the higher-seeded teams tend to choose to play defense in the first round of the SnD series played as the second game, although this choice is not consistent across teams, or with the same team over time.

@@

There have been a handful of studies of the distribution of games played in a series of a professional sport. Most assume a constant probability $\phi$ of a given team winning a game in the series, regardless of the series state. Mosteller [-@mosteller1952] observed that the American League had dominated the National League in MLB World Series matchups through 1952, implying that games should not be modeled with a constant (i.e. the null) $\phi_0 = 0.5$. Mosteller proposed three approaches for identifying the optimal constant probability value of the stronger team in the World Series, finding $\phi \approx 0.65$ in each case: (1) solving for $\phi$ from the observed average number of games won by the loser of the series, which he called the "method of moments" approach; (2) maximizing the likelihood that the sample would have been drawn from a population in which the probability of a team winning a game is constant across the series (i.e. maximum likelihood), and (3) minimizing the chi-squared goodness of fit statistic $\chi^2$ as a function of $\phi$.

Chance [-@chance2020] re-examines the constant probability notion in the MLB World Series (1923--2018), the NBA Finals (1951--2018), and the NHL Stanley Cup (1939--2018). Chance found strong evidence against the null hypothesis of $\phi_0 = 0.5$ in the MLB and NHL championship series when applying Mosteller's first and second methods.

Momentum, one of most discussed topics in sports analytics, goes hand-in-hand with a discussion of the nature of series outcomes.[^13] Two opposing fallacies are observed in the context of momentum: the "gambler's fallacy" (negative recency) and the "hot hand fallacy" (positive recency). Per Ayton et al. [-@ayton2004], negative recency is "the belief that, for random events, runs of a particular outcome ... will be balanced by a tendency for the opposite outcome", while positive recency is the expectation of observing future results that match recent results.

[^13]: We often use "streaks" and momentum interchangeably, but as Steeger et al. ([-@steeger2021]) note, momentum implies dependence between events, whereas streaking does not.

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985], henceforth GVT, ido not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018], henceforth MS, provided a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy. Specifically, MS say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to GVT's data, MS came to the opposite conclusions as GVT.

Other research has borrowed a common approach from the field of quality control, looking at unlikely sequences of events with the Wald-Wolfowitz runs test (Peel and Clauset 2015, Steeger et al. 2021). Peel and Clauset found no evidence for unlikely sequences of scoring events in the NHL, College Football (CFB), and National Football League (NFL), but did in the NBA. As a check on their entropy approach to momentum identification, Steeger et al. found several NHL teams with sequence of wins in the 2018-2019 regular season violating the Wald-Wolfowitz null hypothesis.

@@

Despite the plethora of existing research on games played in a series and momentum in sports, these topics have yet to be investigated heavily, if at all, in esports. Work has been done to examine intra-round win probability in other FPS titles such as Counter-Strike [@xenopoulos2022] and Valorant [@derover2021], both of which are round-based like CoD SnD. However, considering Counter-Strike and Valorant specifically, research on round-level trends seems non-existent, perhaps for one of several reasons:

1.  Both have economic aspects that can create clear advantages for one team in a round, given how prior rounds played out. CoD has no such equivalent, except for perhaps "kill streaks", which infrequently occur.
2.  Teams play either offense or defense for many consecutive rounds. (The former has a 15-15-1 format and the latter uses a 12-12-1 format.) On the other hand, teams in CoD SnD rotate roles every round, analogous to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series for professional sports like the MLB, NBA, and NHL.[^14] While theoretically one might be able to account for any kind of format, such as a 5-5-1, the rotation of team sides every round is convenient for convincing ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli trials.
3.  Both Counter-Strike and Valorant have overtime rules---team must win by two rounds---which can make end-of-series sequences difficult to model. CoD SnD does not have overtime rules.

[^14]: 1-1-1-1-1-1-1 is not used today in these leagues, but it was at least once in each league.

To our knowledge, there is no existing public statistical research on the CoD SnD format, beyond descriptive analysis on social media. While intra-round trends may be more directly applicable to teams looking for an advantage on their competition, broader investigation of a concept like the hot-hand fallacy in a different sport---particularly one that is less subjected to factors that may be difficult to control for, e.g. weather---should be useful as a precedent or supplement for future researchers.

@@

In a best-of-$s$ format, assuming a constant round win probability $\phi$[^15] (where $0 \leq \phi \leq 1$) for one team[^16], the expected proportion of series ending in $r$ rounds ($r \leq s$) is given by Equation \ref{eq:series-length}.

[^15]: If $\phi > 0.5$, we can say that this is the stronger team.

[^16]: $\phi$ and other symbols are selected so as to reserve symbols like $p$ for usage in other contexts without causing confusion for the reader. Upper-case symbol, e.g. $\Phi$, are consistently used in this paper to represent proportions, while lower-case symbols corresponding to upper-case symbols, e.g. $\phi$, are used to represent Bernoulli trial success rates. "Hats", e.g. $\hat{\Phi}$, are used to convey expectations, while bare symbols symbols convey observational data.

```{=tex}
\begin{equation}\label{eq:m}
m = \frac{s + 1}{2}.
\end{equation}
```
```{=tex}
\begin{equation}\label{eq:series-length}
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(s - r)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m).
\end{equation}
```
For example, assuming $\phi = 0.5$, the probability of a series ending in nine rounds in CoD SnD, where $s=11$, is

@@

```{=tex}
\begin{equation}\label{eq:chi-squ}
\chi^2 = \sum^{r \in R} \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)},
\end{equation}
```
in which $R = [6, 7, 8, 9, 10, 11]$ for CoD SnD.

@@

```{=tex}
\begin{equation}\label{eq:pwr}
p_0(\text{win} | r) = p^{+|r}_0 = \begin{cases} 
\frac{m}{r} & \text{team wins series}, \\ 
\frac{r - m}{r} & \text{team loses series},
\end{cases}
\end{equation}
```
using $m$ from Equation \ref{eq:m}.

As shown by MS with their Theorem 1, we should expect the proportion of round wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{+|k,r}_{MS}$, to be strictly less than $p^{+|r}$.[^17] Unfortunately, there does not exist a closed form representation for the expected value of $\hat{P}^{+|k,r}_{MS}$ for $k > 1$. Nonetheless, one may run simulations to estimate the expected value, as we choose to do.[^18]

[^17]: See Appendix E.1 of MS for the proof.

[^18]: We have adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

Although one might be tempted to mirror the hypothesis testing for difference of proportions performed by MS (and GVT), our context is fundamentally different from that of MS, who focus on longitudinal data in controlled settings.[^19] The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series. The Bernoulli trial success probability, i.e. the single round win rate in CoD SnD, is not independent of the opponent. Consequently, a statistical test of the difference in $\hat{P}^{+|k,r}$ and $\hat{P}^{-|kr}$, as performed by MS to evaluate their hypothesis regarding post-streak success rate, is not completely appropriate, although useful as a reference.

[^19]: GVT also perform statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team.

@@

```{=tex}
\begin{equation}\label{eq:pwkr}
p_0(\text{win} | k, r) = p^{+|k,r}_0 = \begin{cases}
\frac{m - k}{r - k} & \text{team wins series}, \\
\frac{s - m - k}{r - k} & \text{team loses series},
\end{cases}
\end{equation}
```
using $m$ from Equation \ref{eq:m}.

We can perform a binomial test to evaluate the null hypothesis $\omega = \omega_0$ for the observed probability of success $\omega$ and a user-specified $\omega_0$ (where $0 \leq \omega_0 \leq 1$). If there are $r^+$ observed successes in a sample of $r$ trials and we expect that there should be $r * \omega_0$, the probability of arriving at this expected number of successes is

```{=tex}
\begin{equation}\label{eq:binom}
\Pr(r^w) = {\binom {r}{r^+}} \omega^{r^+}(1-\omega)^{r-r^+}.
\end{equation}
```
Treating the notional proportion $\hat{P}^{+|k,r}_0$ as the null $\omega_0$ and plugging in the observed proportion $P^{+|k,r}$ for $\omega$ in Equation \ref{eq:binom} (treating proportions as probabilities), we can evaluate the null hypothesis that the observed probability is equal to the notional probability. If we can reject this null hypothesis, then we can consider team momentum, represented by post-streak success, plausible.

One can perform the same binomial test for the MS's streak-selection-adjusted proportion, $\hat{P}^{+|k,r}_{MS}$.[^20]

[^20]: Given the caveats mentioned before, these results should be heeded with caution.

We can further decompose Equation \ref{eq:pwkr} by the round $i$ (where $i \leq r$) in which the streak of length $k$ carries into.

```{=tex}
\begin{equation}\label{eq:pwkri}
p_0(\text{win} | k, r, i) = p^{+|k,r,i}_0 = \begin{cases}
\frac{m - k}{r - k}, & \text{team wins series}, i \neq r, \\
1, & \text{team wins series}, i = r, \\
\frac{s - m - k}{r - k}, & \text{team loses series}, i \neq r, \\
0, & \text{team loses series}, i = r.
\end{cases}
\end{equation}
```
Again, we caapply a binomial test to evaluate the hypothesis that the expected proportions, $\hat{P}^{+|k,r,i}_0$ and $\hat{P}^{+|k,r,i}_{MS}$ separately, are equal to the observed proportion $P^{+|k,r}$.
